BOLD Parameter Estimation Using Particle Filters

### 2010-08-15 doc/thesis/results.tex
more work on the results

To begin the single voxel simulation; I generated a signal using the following parameters:
$\{\tau_0 = 1.45, \alpha = 0.3, E_0 = 0.47, V_0 = 0.044, \tau_s = 1.94, \tau_f = 1.99, \epsilon = 1.8\}$.
These same parameters were used for all the tests in this chapter. The noise realizations were
then generated based on a measurement noise ($\sigma_y$) of $0.001$ and a drift standard deviation
($\sigma_x$) of $0.0005$. The measurement noise as well as the steps of the drift
were taken to be Gaussian.
For this low noise case, the eleven realizations are shown in \autoref{fig:LowNoiseRealization}.
The actual signal delivered into the particle filter
was the result of preprocessing to remove drift, as described in 
\autoref{sec:Methods Preprocessing}. The bias introduced into the signal by preprocessing 
could have some effect on the resulting fit; thus the preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
The data shows that the spline struggles to match the signal near the end, although 
overall, the preprocessing seems to have been successful at removing trends.  In this case
the cause of the problem was the final stimuli that pushed the final median well above the base
level. The estimated BOLD signal is compared to the true BOLD signal in \autoref{fig:FitComparisonLowNoise}.
At several time points the particle filter successfully filtered the input. 
For instance, in the last 30 seconds
the estimates stay flat in spite of the preprocessed data drifting off. By
this point, the algorithm has converged sufficiently to prevent inexplicable movement.
A similar circumstance occurs at 100 seconds in. Note that bias introduced by pre-processing 
results in a slightly larger peak signal. The final parameter sets are shown in 
\autoref{tab:LowNoiseResults}. Note that throughout the results, unless otherwise specified,
the term \emph{residual} will be defined as the square root of the mean squared residual, 
\begin{equation}
\text{Residual} = \sqrt{\sum (\hat{y}_k - y_k)^2}
\end{equation}
where $\hat{y}_k$ is the estimated output at time $k$ and $y_k$ is the preprocessed output
sampled at time $k$. Note the subtle distinction between this and what will be 
termed the \emph{error} which is defines as 
\begin{equation}
\text{Error} = \sqrt{\sum (\hat{y}_k - Y_k)^2}
\end{equation}
where $\hat{y}_k$ is the estimated output, and $Y_k$ is the underlying (free of noise)
signal. Throughout this section and the next, the terms error and residual
will be referencing these values.

@@

For the high noise simulation, the exact same procedure was followed as in \autoref{sec:SimLowNoise}
except that $\sigma_y$ and $\sigma_x$ were set to $0.01$ and $0.005$,
respectively. This is an order of magnitude higher than the previous tests, and indeed the
noise appears to dominate the output, as seen in \autoref{fig:HighNoiseRealization}.  
As before, the results of preprocessing are helpful in identifying the reason for
bias in the estimated output signal; thus the preprocessed signals are shown in 
\autoref{fig:PreprocessedHighNoise}. The results of the particle filter
for each of the eleven runs are shown in \autoref{fig:FitComparisonHighNoise}. 
The preprocessing again led the estimate
to somewhat higher activation levels, and it would appear that the subtleties of
different time constants are being lost due to the noise.

@@

Two runs are shown in more detail in \autoref{fig:NoiseComparisonJustTwo}.
Run 1 has higher peaks, in spite of the fact that both signals are based
on the same underlying signal. There also appears
to be more drift than the 20 measurements per knot could fit, which explains run 1's
prolonged increase at 170 seconds. Interestingly, run 1 and run 2 emphasize
different aspects of the signal. Run 2 had a much better match to the 
peaks, when compared to the true signal, yet run 1 matched the post-stimulus
undershoot better. Its worth noting that the post-stimulus undershoots
here are much shorter than the observed prolonged post-stimulus undershoot
discussed in \autoref{sec:Post Stimulus Undershoot}. 
Despite the rather significant amount of noise, the ultimate results are 
actually quite good. \autoref{tab:HighNoiseResults} shows
the error  for all eleven runs, and highlights the two runs analyzed 
in \autoref{fig:ConvergenceRuns1} and \autoref{fig:ConvergenceRuns2}.

\begin{figure}
\subfigure[$\tau_0$, $\alpha$, $E_0$, $V_0$, Run 1]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run5_1}}\\
\end{figure}

\begin{figure}
\subfigure[$\tau_s$, $\tau_f$, $\epsilon$, $v$, Run 1] 
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run5_2}}\\
\end{figure}

\begin{figure}
\subfigure[$q$, $s$, $f$, $BOLD$, Run 1 ]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run5_3}}
\caption{Converging histogram for parameters during run 1, as in \autoref{fig:NoiseComparisonJustTwo}.}
\label{fig:ConvergenceRuns1}
\end{figure}

\begin{figure}
\subfigure[$\tau_0$, $\alpha$, $E_0$, $V_0$, Run 2]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run6_1}}\\
\end{figure}
\begin{figure}
\subfigure[$\tau_s$, $\tau_f$, $\epsilon$, $v$, Run 2] 
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run6_2}}\\
\end{figure}
\begin{figure}
\subfigure[$q$, $s$, $f$, $BOLD$, Run 2 ]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run6_3}}
\caption{Converging histogram for parameters during run 2, as in \autoref{fig:NoiseComparisonJustTwo}.}
\label{fig:ConvergenceRuns2}
\end{figure}

There are a number of interesting convergence properties of the
particle filter when more noise is present, as both \autoref{fig:ConvergenceRuns1} and
\autoref{fig:ConvergenceRuns2} show. The particle filter seems to converge
significantly faster; as points tend to be further out on the weighting function. This
also causes significantly more resampling which is the explanation for the percieved
jumps in resolution that occur from time to time. Here it is clear that the mode and
the mean will not be substantially different. Because of the increased noise, its likely
that the weighting function is not sufficiently wide to account for the measurement noise.

The parameters arrived at for all ten filter runs are shown in \autoref{tab:HighNoiseResults}.
Clearly the additional noise have resulted in much more sporadic results. This is often the 
result when the particle filter converges too fast, a result of the weighting functions' variance
being smaller than the measurement noise ($0.005$ vs. $0.01$). The error
certainly suffers due to this effect.

%NO SIGNAL, LOW NOISE
\subsection{Pure-Noise, low magnitude}
\label{sec:PureNoiseLowMag}
The next two single-voxel tests force the particle filter to attempt to learn a noise-only
time series. In this first test the noise used is the same as that from the \autoref{sec:SimHighNoise},
$\sigma_x = 0.01, \sigma_y = 0.005$. The parameters will be set the same as well, but the
stimulus/input neuronal efficiency ($\epsilon$) will be set to 0. In effect it is a region of the brain for
which no visual stimulus induces activity. The question then is how the
particle filter will respond to signal that does not correlate with the input, and how
the output may be differentiated from a time series that does. The time series'
are shown in \autoref{fig:NoiseOnly}, the preprocessed versions are shown in \autoref{fig:PreprocessedNoiseOnly}.
The line fits are shown in \autoref{fig:fits_noiseonly}.

@@

The data shows that the parameters have not taken any visible turns into
unrealistic territory; individual parameters will not be a distinguishing factor
in determining the validity of the model. Additionally the residual 
also appears to be lower than the residual of the low noise simulation from earlier. 
The precision with which the convergence occurred is actually the best seen in any
simulation.  The reason for the low error is that the overall
signal is significantly smaller than any other simulation in this thesis.
The peaks never even 1\% difference, with most staying even below .5\% 
(\autoref{fig:PreprocessedNoiseOnly}). Because the stayed 
well within the range of $0.005$, which is the standard deviation of the weighting function
the particles did not ever get substantially different weights. Thus, the reason
the fits are in such agreement is that no particles gained increased weight, and so
no convergence occurred at all, as further shown by \autoref{fig:JustNoiseConvergence}. 
This is good evidence that a per time-series weighting function could benefit
even out the convergence rate; and thus allow better determination between signals
like this and low noise case from earlier.
Thus we have now seen how the particle filter responds when the signal is completely
random and low level, however there is also some possibility that the signal will
be random, yet active. Thus one more test is necessary; time series with no 
underlying signal (as far as responding stimuli) but with enough noise to 
actually be considered a signal.

>>>>

To begin the single voxel simulation; I generated a signal using the following parameters:
$\{\tau_0 = 1.45, \alpha = 0.3, E_0 = 0.47, V_0 = 0.044, \tau_s = 1.94, \tau_f = 1.99, \epsilon = 1.8\}$.
These same parameters were used for all the tests in this chapter (except when the signal
consists entirely of noise). The noise realizations were
then generated based on a measurement noise ($\sigma_y$) of $0.001$ and a drift standard deviation
($\sigma_x$) of $0.0005$. The measurement noise as well as the steps of the drift
were taken to be Gaussian.
For this low noise case, the eleven realizations are shown in \autoref{fig:LowNoiseRealization}.
The actual signal delivered into the particle filter
was the result of preprocessing to remove drift, as described in 
\autoref{sec:Methods Preprocessing}. The bias introduced into the signal by preprocessing 
could have some effect on the resulting fit; thus the preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
The data shows that the spline struggles to match the signal near the end, although 
overall, the preprocessing seems to have been successful at removing trends.  In this case
the cause of the problem was the final stimuli that pushed the final median well above the base
level. The estimated BOLD signal is compared to the true BOLD signal in \autoref{fig:FitComparisonLowNoise}.
At several time points the particle filter successfully filtered the input. 
For instance, in the last 30 seconds
the estimates stay flat in spite of the preprocessed data drifting off. By
this point, the algorithm has converged sufficiently to prevent inexplicable movement.
A similar circumstance occurs at 100 seconds in. Note that bias introduced by pre-processing 
results in a slightly larger peak signal. The final parameter sets are shown in 
\autoref{tab:LowNoiseResults}. Note that throughout the results, unless otherwise specified,
the term \emph{residual} will be defined as the square root of the mean squared residual, 
\begin{equation}
\text{Residual} = \sqrt{\sum (\hat{y}_k - y_k)^2}
\end{equation}
where $\hat{y}_k$ is the estimated output at time $k$ and $y_k$ is the preprocessed output
sampled at time $k$. Note the subtle distinction between this and what will be 
termed the \emph{error} which is defines as 
\begin{equation}
\text{Error} = \sqrt{\sum (\hat{y}_k - Y_k)^2}
\end{equation}
where $\hat{y}_k$ is the estimated output, and $Y_k$ is the underlying (free of noise)
signal. Throughout this section and the next, the terms error and residual
will be referencing these values.

@@

For the high noise simulation, the exact same procedure was followed as in \autoref{sec:SimLowNoise}
except that $\sigma_y$ and $\sigma_x$ were set to $0.01$ and $0.005$,
respectively. This is an order of magnitude higher than the previous tests, and indeed the
noise appears to dominate the output, as \autoref{fig:HighNoiseRealization} shows. 
The preprocessed signals are shown in 
\autoref{fig:PreprocessedHighNoise} and the results of the particle filter
for each of the eleven runs are shown in \autoref{fig:FitComparisonHighNoise}. 
The noise and preprocessing again led the estimate
to somewhat higher activation levels, and it would appear that the subtleties of
different time constants are being lost in the noise.

@@

Two runs are shown in more detail in \autoref{fig:NoiseComparisonJustTwo}.
Run 1 has higher peaks, in spite of the fact that both signals are based
on the same underlying signal. There also appears
to be more drift than the 20 measurements per knot could fit, which explains run 1's
prolonged increase at 170 seconds. Interestingly, run 1 and run 2 emphasize
different aspects of the signal. Run 2 had a much better match to the 
peaks, when compared to the true signal, yet run 1 matched the post-stimulus
undershoot better. Despite the rather significant amount of noise, the ultimate results are 
actually quite good. \autoref{tab:HighNoiseResults} shows
the error  for all eleven runs, and highlights the two runs analyzed 
in \autoref{fig:ConvergenceRuns1} and \autoref{fig:ConvergenceRuns2}.

\begin{figure}[H]
\subfigure[$\tau_0$, $\alpha$, $E_0$, $V_0$, Run 1]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run5_1}}\\
\end{figure}

\begin{figure}[H]
\subfigure[$\tau_s$, $\tau_f$, $\epsilon$, $v$, Run 1] 
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run5_2}}\\
\end{figure}

\begin{figure}[H]
\subfigure[$q$, $s$, $f$, $BOLD$, Run 1 ]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run5_3}}
\caption{Converging histogram for parameters during run 1, as in \autoref{fig:NoiseComparisonJustTwo}.}
\label{fig:ConvergenceRuns1}
\end{figure}

\begin{figure}[H]
\subfigure[$\tau_0$, $\alpha$, $E_0$, $V_0$, Run 2]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run6_1}}\\
\end{figure}

\begin{figure}[H]
\subfigure[$\tau_s$, $\tau_f$, $\epsilon$, $v$, Run 2] 
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run6_2}}\\
\end{figure}

\begin{figure}[H]
\subfigure[$q$, $s$, $f$, $BOLD$, Run 2 ]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run6_3}}
\caption{Converging histogram for parameters during run 2, as in \autoref{fig:NoiseComparisonJustTwo}.}
\label{fig:ConvergenceRuns2}
\end{figure}

The particles converged much faster when more noise was present (\autoref{fig:ConvergenceRuns1}, 
\autoref{fig:ConvergenceRuns2}). This also caused significantly more resampling which 
is the explanation for the percieved jumps in the histograms. 
The parameters arrived at for all 11 runs are shown in \autoref{tab:HighNoiseResults}.
Clearly the additional noise has resulted in much more sporadic results. This is often the 
result when the particle filter converges too fast, in this case the result of the 
weighting functions' variance being smaller than the measurement noise ($0.005$ vs. $0.01$). 
The error suffers due to this effect.

%NO SIGNAL, LOW NOISE
\subsection{Pure Noise, Low magnitude}
\label{sec:PureNoiseLowMag}
The next two single-voxel tests forced the particle filter to attempt to learn a noise-only
time series. In this first test the noise used is the same as that from the \autoref{sec:SimHighNoise},
$\sigma_x = 0.01, \sigma_y = 0.005$. The parameters were set the same as well, but the
stimulus neuronal efficiency ($\epsilon$) was set to 0. In effect it simulates a region 
of the brain that does respond to the stimulus. The importance of this test is
to determine how the
particle filter responds to a signal that is not correlated with the input. Additionally,
this test determines how the output may be differentiated from the output of the previous
two sections. The time series'
are shown in \autoref{fig:NoiseOnly} and the preprocessed versions are shown 
in \autoref{fig:PreprocessedNoiseOnly}.
The line fits are shown in \autoref{fig:fits_noiseonly}.

@@

The data shows that the parameters did not converge to any unreasonable
values; individual parameters could not be used as a distinguishing factor
to determine the validity of the model. Additionally the residual 
also appears to be lower than the residual of the low noise simulation from \autoref{sec:SimLowNoise}. 
The precision with which the convergence occurred is actually the best seen in any
simulation.  The reason for the low error is that the overall
signal is significantly smaller than any other simulation in this thesis.
The peaks never even 1\% difference, with most staying even below .5\% 
(\autoref{fig:PreprocessedNoiseOnly}). Because the stayed 
well within the range of $0.005$, which is the standard deviation of the weighting function
the particles did not ever get substantially different weights. Thus, the reason
the fits are in such agreement is that no particles gained increased weight, and so
no convergence occurred at all, as further shown by \autoref{fig:JustNoiseConvergence}. 
This is good evidence that a per time-series weighting function could benefit
even out the convergence rate; and thus allow better determination between signals
like this and low noise case from earlier.
Thus we have now seen how the particle filter responds when the signal is completely
random and low level, however there is also some possibility that the signal will
be random, yet active. Thus one more test is necessary; time series with no 
underlying signal (as far as responding stimuli) but with enough noise to 
actually be considered a signal.
